Left-linear rules in adhesive categories

### 2024-02-13 icalp/fusione.tex
3 passi

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two consecutive steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  those steps could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting. More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, the equivalence 
captures when 
%sequential 
two steps are causally unrelated and thus could be executed in any 
order, and possibly in parallel. 
%
Seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
%
Concerning formalisms based on the rewriting paradigm, i.e. where the
behaviour is dictated by a set of rules that have to be instantiated with
the system at hand, concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
Noteworthy examples are
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

@@

There is, however, a further problem. In the literature on linear
rewriting systems, a central result is the a characterisation of
switch equivalence of sequences of steps based on what are 
called processes or consistent permutations~\cite{Handbook,tobias}.
%\todo{Handbook, graph processes, tobias}
In turn this is the key
to derive some simple but fundamental properties for linear rewriting
systems, namely
\begin{enumerate}
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

@@

\propUnique*

\label{propUnique-proof}
\begin{proof}
  Both pairs $(k, f)$ and $(k', f')$ are pushout complements of $l$
  and $n$, thus, \Cref{lem:pocomp} yields an isomorphism
  $\phi_D\colon D\to D'$. Computing we have
  \[
    g'\circ \phi_D \circ k= g' \circ k'=h'\circ r
  \]

  \noindent
  \parbox{3cm}{
    $\xymatrix{K \ar@/^.4cm/[rr]^{k'}\ar[d]_{r} \ar[r]_{k}&
      \ar[r]_{\phi_D} D \ar[d]^{g}& D' \ar[d]^{g'}\\ R
      \ar@/_.4cm/[rr]_{h'} \ar[r]^{h}& H \ar[r]^{\phi_H}& H'}$}
  \hfill
  \parbox{10cm}{ \hspace{15pt}Hence, we have another
    $\phi_H\colon H\to H'$. To see that $\phi_H$ is an isomorphism,
    consider the diagram aside. By hypothesis the whole rectangle and
    its left half are pushouts, therefore, by \Cref{lem:po1} its right
    square is a pushout too. The claim now follows from the fact that
    the pushout of an isomorphism is an isomorphism. \qedhere }
\end{proof}

@@

\begin{proof}
    By definition of switch we know that $\der{E}_0$ and $\der{F}_0$ have the same match $f_0\circ i_1$ and that $\der{E}_1$ and $\der{F}_1$ have the same comatch $g_1\circ i_0$. Thus we can build the solid part of the diagram below.
    \[\xymatrix@C=22pt{G_{0} && \ar@{>->}[ll]_{f_{\der{F},0}} D_{\der{F},0} \ar[rr]^{g_{\der{F},0}}&& G_{\der{F},1} && \ar@{>->}[ll]_{f_{\der{F},1}} D_{\der{F},1} \ar[rr]^{g_{\der{F},1}}&& G_{2}\\L_1 \ar[d]^{f_0\circ i_1} \ar[u]_{f_0\circ i_1}&& K_{1} \ar[d]^{k_{\der{E},0}} \ar[u]_{k_{\der{F},0}}\ar@{>->}[ll]_{l_1} \ar[r]^{r_1} & R_{1} \ar@/^.35cm/[drrr]|(.3)\hole_(.4){j_0} \ar[dr]|(.3)\hole_{h_{\der{E},0}} \ar@/_.35cm/[urrr]|(.3)\hole^(.4){a_0} \ar[ur]|(.3)\hole^{h_{\der{F},0}}& &L_{0} \ar@/_.35cm/[dlll]^(.4){j_1} \ar[dl]|(.3)\hole^{m_{\der{E},1}} \ar@/^.35cm/[ulll]_(.4){a_1} \ar[ul]|(.3)\hole_{m_{\der{F},1}}& K_{0} \ar[d]_{k_{\der{E},1}} \ar[u]^{k_{\der{F},1}}\ar@{>->}[l]_{l_{0}} \ar[rr]^{r_0} && R_{0} \ar[d]_{g_1\circ i_0} \ar[u]^{g_1\circ i_0}\\ \ar@/^.5cm/[uu]^{\id{G_0}} G_{0} && \ar[ll]^{f_{\der{E},0}} D_{\der{E},0} \ar@{.>}@/^.5cm/[uu]^(.33){\phi_{0}}|\hole \ar[rr]_{g_{\der{E},0}}&& G_{\der{E},1} \ar@{.>}[uu]^(.76){\psi_1}|(.45)\hole |(.55)\hole&& \ar@{>->}[ll]^{f_{\der{E},1}} D_{\der{E},1} \ar@{.>}@/_.5cm/[uu]_(.33){\phi_{1}}|\hole\ar[rr]_{g_{\der{E},1}}&& G_{2}\ar@{.>}@/_.5cm/[uu]_{\psi_2}}\]

@@

    Now, the square
    \[\xymatrix@C=30pt{K_0 \ar[d]_{k_{\der{E,1}}} \ar@{>->}[r]^{l_0}& L_0 \ar[d]^{\psi_1\circ m_{\der{E},1}} \\D_{\der{E},1} \ar@{>->}[r]_{\psi_1\circ f_{\der{E},1}}& G_{\der{F}, 1}} \]
    is a pushout and $\psi_1\circ m_{\der{E},1}=m_{\der{F},1}$, thus, by \Cref{prop:unique} there are $\phi_1\colon D_{\der{E},1}\to D_{\der{F},1}$ and $\psi_2\colon G_2\to G_2$ fitting in the diagram below
    \[\xymatrix{G_{\der{F},1} & \ar@{>->}[l]_{f_{\der{F},1}} D_{\der{F},1} \ar[r]^{g_{\der{F},1}}& G_2 \\L_0 \ar[d]^{m_{\der{E},1}} \ar[u]_{m_{\der{F},1}}& K_{1} \ar[d]^{m_{\der{E},1}} \ar[u]_{m_{\der{F},1}}\ar[l]_{l_1} \ar[r]^{r_1} & R_{1}  \ar[d]_{g_1\circ i_0}  \ar[u]^{g_1\circ i_0}\\ \ar@/^.5cm/[uu]^{\psi_1} G_{\der{E},1} & \ar@{>->}[l]^{f_{\der{E},1}} D_{\der{E},1} \ar@/^.5cm/[uu]^(.33){\phi_{1}}|\hole \ar[r]_{g_{\der{E},1}}& G_2 \ar@/_.5cm/[uu]_{\psi_2}}\]
    To conclude, it is now enough to notice that
    \[f_{\der{F},1} \circ \phi_1\circ j_0=\psi_1\circ f_{\der{E},1}\circ j_0=\psi_1\circ h_{\der{E,0}}=h_{\der{F},0}=f_{\der{F},1}\circ a_0\]
    allowing us to deduce $ \phi_1\circ j_0=a_0$.
    \qedhere
\end{proof}

\begin{proposition}
	\label{prop:tec}
	Let $(i_0, i_1)$ be an independence pair between two direct
	derivations $\dder{D}_0$ and $\dder{D}_1$ in a left-linear
	DPO rewriting system $(\X,\R)$. Consider the pullback of
	$f_1\colon D_1\mto G_1$ along $g_0\colon D_0\to G_1$ (first
	square below), then there exist the arrows
	$u_1\colon K_1\to P$ and $u_0\colon K_0\to P$ fitting in the
	central and right square. Moreover, the right one is always
	a pullback.
	\[\xymatrix{P \ar@{>->}[r]^{p_0} \ar[d]_{p_1} & D_0
		\ar[d]^{g_0}& K_0 \ar[r]^{r_0} \ar@{.>}[d]_{u_0}& R_0
		\ar[d]^{i_0} & K_1 \ar@{>->}[r]^{l_1} \ar@{.>}[d]_{u_1}&
		L_1 \ar[d]^{i_1}\\ D_1 \ar@{>->}[r]_{f_1} &G_1 &P
		\ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0}
	\]
\end{proposition}

\begin{proof}
	We start noticing that
	\[
	f_1\circ i_0\circ r_0=h_0\circ r_0=g_0\circ k_0 \qquad
	g_0\circ i_1\circ l_1=m_1\circ l_1= f_1 \circ k_1
	\]
	Thus there exists the dotted $u_0\colon K_0\to P$,
	$u_1\colon K_1\to P$ in the diagram
	\[\xymatrix{K_0 \ar[d]_{r_0}
		\ar@{.>}[r]_{u_0}\ar@/^.4cm/[rr]^{k_0} &P
		\ar@{>->}[r]_{p_0} \ar[d]_{p_1}& D_0 \ar[d]^{g_0}&K_1
		\ar@{>->}[d]_{l_1}
		\ar@{.>}[r]_{u_1}\ar@/^.4cm/[rr]^{k_1} &P \ar[r]_{p_1}
		\ar@{>->}[d]_{p_0}& D_1 \ar@{>->}[d]^{f_1}\\R_0
		\ar@/_.4cm/[rr]_{h_0}\ar[r]^{i_0}& D_1
		\ar@{>->}[r]^{f_1}& G_1&L_1
		\ar@/_.4cm/[rr]_{m_1}\ar[r]^{i_1}& D_0 \ar[r]^{g_0}&
		G_1}\]
	By \Cref{prop:pbpoad,lem:pb1} we get that the left half of
	the second rectangle is a pullback.
	\qedhere
\end{proof}

@@

\begin{proof}
  Consider the arrow $u_1\colon K_1\to P$ obtained using
  \Cref{prop:tec}. It fits in the diagram:
\[		\xymatrix{K_1 \ar[r]^{u_1} \ar@{>->}[d]_{l_1}&P
	\ar[r]^{\id{P}}\ar@{>->}[d]_{p_0}& P\ar@{>->}[d]^{p_0}\\L_1
	\ar[d]_{\id{L_1}}\ar[r]^{i_1}&D_0 \ar[r]^{\id{D_0}}
	\ar[d]_{\id{D_0}}& D_{0} \ar@{>->}[d]^{f_0}\\
	L_1\ar[r]^{i_1} \ar@/_.4cm/[rr]_{m_{\der{E},0}}&D_0
	\ar@{>->}[r]^{f_0}&G_0}\]

\smallskip \noindent 
  \parbox{4cm}{$\xymatrix{K_1 \ar@{>->}[d]_{l_1}\ar[r]^{k_{\der{E},0}}&
  		D_{\der{E},0} \ar@{>->}[d]^{f_{\der{E},0}} & K_1 \ar[r]^{u}
  		\ar@{>->}[d]_{l_1}& P \ar@{>->}[d]_{f_0\circ
  			p_0}\\L_1\ar[r]_{m_{\der{E},0}}&G_0 &L_1
  		\ar[r]_{m_{\der{E},0}}&G_0} $}\hspace{2cm}
  	\parbox{8cm}{\hspace{15pt} Since $f_0$ is a monomorphism, every square in the above diagram
  is a pullback. Thus the whole right square is a pullback too.
  Applying \Cref{lem:radj} to the pair of squares on the left now gives
  us the wanted $q_0\colon P\to D_{\der{E},0}$. \qedhere}

@@

	Moreover, by \Cref{prop:pbpoad}, the first square below is a
	pullback, therefore, by \cref{rem:deco} the last three squares
	witness that $(j_0, j_1)$ is a strong independence pair too.
	\[
	\xymatrix{P\ar[r]^{q_1} \ar@{>->}[d]_{q_0}&Q_1
		\ar@{>->}[d]^{a_1}&K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1
		\ar[d]^{j_0} & K_0 \ar@{>->}[r]^{l_0} \ar[d]_{u_0}& L_0
		\ar[d]^{j_1}&K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0\ar[d]^{i_0} \\
		Q_ 0\ar[r]_{b_0}&H_1&P \ar[r]_{q_1} & Q_1 &P \ar[r]_{q_0} &
		Q_0&P \ar[r]_{p_1} & D_1}
	\]
\end{remark}

@@

\begin{proof}
	To fix the notation, let us depict the$\der{D}=\{\dder{D}_i\}_{i=0}^2$:
	\[\xymatrix@C=15pt{L_0\ar[d]_{m_0}&& K_0 \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0 \ar@/^.35cm/[drrr]_(.4){i_0}|(.285)\hole \ar[dr]|(.28)\hole_{h_0} && L_1 \ar@/_.35cm/[dlll]^(.4){i_1} \ar[dl]|(.28)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[r]^{r_1} & R_1 \ar[dr]|(.28)\hole_{h_1} \ar@/^.35cm/[drrr]_(.4){j_0}|(.285)\hole  && L_2 \ar@/_.35cm/[dlll]^(.4){j_1} \ar[dl]|(.28)\hole^{m_2}& K_2 \ar[d]^{k_2}\ar@{>->}[l]_{l_2} \ar[rr]^{r_2} && R_2 \ar[d]^{h_2} \\G_0 && \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_{g_0}&& G_1  && \ar@{>->}[ll]^{f_1} D_1 \ar[rr]_{g_1}&& G_2 && \ar@{>->}[ll]^{f_2} D_2 \ar[rr]_{g_2}&& G_3 }\]

@@

	We begin switching $\dder{D}_0$ and $\dder{D}_1$.

	\[\xymatrix@C=15pt{L_1\ar[d]_{f_0\circ i_1}&& K_1 \ar[d]_{k'_1}\ar[ll]_{l_1} \ar[r]^{r_1} & R_1 \ar@/^.35cm/[drrr]_(.4){a_0}|(.285)\hole \ar[dr]|(.28)\hole_{h'_1} && L_0 \ar@/_.35cm/[dlll]^(.4){a_1} \ar[dl]|(.28)\hole^{m'_0}& K_0 \ar[d]^{k'_0}\ar[l]_{l_0} \ar[r]^{r_0} & R_0 \ar[dr]|(.28)\hole_{g_1\circ j_0} \ar@/^.35cm/[drrr]_(.4){b_0}|(.285)\hole  && L_2 \ar@/_.35cm/[dlll]^(.4){b_1} \ar[dl]|(.28)\hole^{m_2}& K_2 \ar[d]^{k_2}\ar[l]_{l_2} \ar[rr]^{r_2} && R_2 \ar[d]^{h_2} \\G_0 && \ar[ll]^{f'_0} D'_0 \ar[rr]_{g'_0}&& G'_1  && \ar[ll]^{f'_1} D'_1 \ar[rr]_{g'_1}&& G_2 && \ar[ll]^{f_2} D_2 \ar[rr]_{g_2}&& G_3 }\]

	Moreover, we know that
	\[m_0=f'_0\circ a_1 \qquad h_1=g'_1\circ a_0\]

>>>>

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting. More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, the equivalence 
captures when 
%sequential 
steps are causally unrelated and thus could be executed in any 
order, and possibly in parallel. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
%
Concerning formalisms based on the rewriting paradigm, i.e. where the
behaviour is dictated by a set of rules that have to be instantiated with
the system at hand, concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
Noteworthy examples are
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

@@

% \begin{enumerate}
% \item \emph{confluence of switching}: when transforming derivation
%   sequences into equivalent ones by switching independent steps, the
%   result not depends on the order in which switchings are applied, but
%   only on the resulting permutation;

% \item \emph{globality of independence}: if two independent steps are
%   moved forward or backward due to the application of switchings to a
%   derivation, they remain independent.
% \end{enumerate}

There is, however, a further problem. In the literature on linear
rewriting systems, a central result is the a characterisation of
switch equivalence of sequences of steps based on what are called processes or
consistent permutations~\cite{Handbook, tobias}.\todo{Handbook, graph processes, tobias}
In turn this is the key
to derive some simple but fundamental properties for linear rewriting
systems, namely

\begin{enumerate}
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

@@

\propUnique*
\label{propUnique-proof}
%\vspace{-.2cm}
\begin{proof}
  Both pairs $(k, f)$ and $(k', f')$ are pushout complements of $l$
  and $n$, thus, \Cref{lem:pocomp} yields an isomorphism
  $\phi_D\colon D\to D'$. An easy computation shows that $g'\circ \phi_D \circ k=h'\circ r$.
  \noindent
  \parbox{3cm}{
    $\xymatrix{K \ar@/^.4cm/[rr]^{k'}\ar[d]_{r} \ar[r]_{k}&
      \ar[r]_{\phi_D} D \ar[d]^{g}& D' \ar[d]^{g'}\\ R
      \ar@/_.4cm/[rr]_{h'} \ar[r]^{h}& H \ar[r]^{\phi_H}& H'}$}
  \hfill
  \parbox{10cm}{ \hspace{15pt}Hence, we have another
    $\phi_H\colon H\to H'$. To see that $\phi_H$ is an isomorphism,
    consider the diagram aside. By hypothesis the whole rectangle and
    its left half are pushouts, therefore, by \Cref{lem:po1} its right
    square is a pushout too. The claim now follows from the fact that
    the pushout of an isomorphism is an isomorphism. \qedhere }
\end{proof}

@@

\begin{proof}
    By definition of switch $\der{E}_0$ and $\der{F}_0$ have the same match $f_0\circ i_1$ and that $\der{E}_1$ and $\der{F}_1$ have the same comatch $g_1\circ i_0$. Thus we get the solid part of the diagram below.
    \[\xymatrix@C=22pt{G_{0} && \ar@{>->}[ll]_{f_{\der{F},0}} D_{\der{F},0} \ar[rr]^{g_{\der{F},0}}&& G_{\der{F},1} && \ar@{>->}[ll]_{f_{\der{F},1}} D_{\der{F},1} \ar[rr]^{g_{\der{F},1}}&& G_{2}\\L_1 \ar[d]^{f_0\circ i_1} \ar[u]_{f_0\circ i_1}&& K_{1} \ar[d]^{k_{\der{E},0}} \ar[u]_{k_{\der{F},0}}\ar@{>->}[ll]_{l_1} \ar[r]^{r_1} & R_{1} \ar@/^.35cm/[drrr]|(.3)\hole_(.4){j_0} \ar[dr]|(.3)\hole_{h_{\der{E},0}} \ar@/_.35cm/[urrr]|(.3)\hole^(.4){a_0} \ar[ur]|(.3)\hole^{h_{\der{F},0}}& &L_{0} \ar@/_.35cm/[dlll]^(.4){j_1} \ar[dl]|(.3)\hole^{m_{\der{E},1}} \ar@/^.35cm/[ulll]_(.4){a_1} \ar[ul]|(.3)\hole_{m_{\der{F},1}}& K_{0} \ar[d]_{k_{\der{E},1}} \ar[u]^{k_{\der{F},1}}\ar@{>->}[l]_{l_{0}} \ar[rr]^{r_0} && R_{0} \ar[d]_{g_1\circ i_0} \ar[u]^{g_1\circ i_0}\\ \ar@/^.5cm/[uu]^{\id{G_0}} G_{0} && \ar[ll]^{f_{\der{E},0}} D_{\der{E},0} \ar@{.>}@/^.5cm/[uu]^(.33){\phi_{0}}|\hole \ar[rr]_{g_{\der{E},0}}&& G_{\der{E},1} \ar@{.>}[uu]^(.76){\psi_1}|(.45)\hole |(.55)\hole&& \ar@{>->}[ll]^{f_{\der{E},1}} D_{\der{E},1} \ar@{.>}@/_.5cm/[uu]_(.33){\phi_{1}}|\hole\ar[rr]_{g_{\der{E},1}}&& G_{2}\ar@{.>}@/_.5cm/[uu]_{\psi_2}}\]

@@

    Now, the square on the left in the diagram below is a pushout and $\psi_1\circ m_{\der{E},1}=m_{\der{F},1}$. Thus, by \Cref{prop:unique} there are the dotted $\phi_1\colon D_{\der{E},1}\to D_{\der{F},1}$ and $\psi_2\colon G_2\to G_2$ in the diagram on the right.
    \[\xymatrix{&&G_{\der{F},1} & \ar@{>->}[l]_{f_{\der{F},1}} D_{\der{F},1} \ar[r]^{g_{\der{F},1}}& G_2 \\K_0 \ar[d]_{k_{\der{E,1}}} \ar@{>->}[r]^{l_0}& L_0 \ar[d]^{\psi_1\circ m_{\der{E},1}}&L_0 \ar[d]^{m_{\der{E},1}} \ar[u]_{m_{\der{F},1}}& K_{1} \ar[d]^{m_{\der{E},1}} \ar[u]_{m_{\der{F},1}}\ar@{>->}[l]_{l_1} \ar[r]^{r_1} & R_{1}  \ar[d]_{g_1\circ i_0}  \ar[u]^{g_1\circ i_0}\\D_{\der{E},1} \ar@{>->}[r]_{\psi_1\circ f_{\der{E},1}}& G_{\der{F}, 1}& \ar@/^.5cm/[uu]^{\psi_1} G_{\der{E},1} & \ar@{>->}[l]^{f_{\der{E},1}} D_{\der{E},1} \ar@{.>}@/^.5cm/[uu]^(.33){\phi_{1}}|\hole \ar[r]_{g_{\der{E},1}}& G_2 \ar@{.>}@/_.5cm/[uu]_{\psi_2}}\]

    To conclude, it is now enough to notice that
    \[f_{\der{F},1} \circ \phi_1\circ j_0=\psi_1\circ f_{\der{E},1}\circ j_0=\psi_1\circ h_{\der{E,0}}=h_{\der{F},0}=f_{\der{F},1}\circ a_0\]
    allowing us to deduce $ \phi_1\circ j_0=a_0$.
    \qedhere
\end{proof}

\begin{proposition}
	\label{prop:tec}
	Let $(i_0, i_1)$ be an independence pair between two direct
	derivations $\dder{D}_0$ and $\dder{D}_1$ in a left-linear
	DPO rewriting system $(\X,\R)$. Consider the pullback of
	$f_1\colon D_1\mto G_1$ along $g_0\colon D_0\to G_1$ (first
	square below), then there exist the arrows
	$u_1\colon K_1\to P$ and $u_0\colon K_0\to P$ fitting in the
	central and right square. Moreover, the right one is always
	a pullback.
	\[\xymatrix@R=15pt{P \ar@{>->}[r]^{p_0} \ar[d]_{p_1} & D_0
		\ar[d]^{g_0}& K_0 \ar[r]^{r_0} \ar@{.>}[d]_{u_0}& R_0
		\ar[d]^{i_0} & K_1 \ar@{>->}[r]^{l_1} \ar@{.>}[d]_{u_1}&
		L_1 \ar[d]^{i_1}\\ D_1 \ar@{>->}[r]_{f_1} &G_1 &P
		\ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0}
	\]
\end{proposition}

\begin{proof}
	We start noticing that
	\[
	f_1\circ i_0\circ r_0=h_0\circ r_0=g_0\circ k_0 \qquad
	g_0\circ i_1\circ l_1=m_1\circ l_1= f_1 \circ k_1
	\]
	Thus there exists the dotted $u_0\colon K_0\to P$,
	$u_1\colon K_1\to P$ in the diagram
	\[\xymatrix@R=15pt{K_0 \ar[d]_{r_0}
		\ar@{.>}[r]_{u_0}\ar@/^.4cm/[rr]^{k_0} &P
		\ar@{>->}[r]_{p_0} \ar[d]_{p_1}& D_0 \ar[d]^{g_0}&K_1
		\ar@{>->}[d]_{l_1}
		\ar@{.>}[r]_{u_1}\ar@/^.4cm/[rr]^{k_1} &P \ar[r]_{p_1}
		\ar@{>->}[d]_{p_0}& D_1 \ar@{>->}[d]^{f_1}\\R_0
		\ar@/_.4cm/[rr]_{h_0}\ar[r]^{i_0}& D_1
		\ar@{>->}[r]^{f_1}& G_1&L_1
		\ar@/_.4cm/[rr]_{m_1}\ar[r]^{i_1}& D_0 \ar[r]^{g_0}&
		G_1}\]
	By \Cref{prop:pbpoad,lem:pb1} we get that the left half of
	the second rectangle is a pullback.
	\qedhere
\end{proof}

@@

\begin{proof}
  Consider the arrow $u_1\colon K_1\to P$ obtained using
  \Cref{prop:tec}. It fits in the big square on the left:
\[		\xymatrix@R=15pt{K_1 \ar[r]^{u_1} \ar@{>->}[d]_{l_1}&P
	\ar[r]^{\id{P}}\ar@{>->}[d]_{p_0}& P\ar@{>->}[d]^{p_0}\\L_1
	\ar[d]_{\id{L_1}}\ar[r]^{i_1}&D_0 \ar[r]^{\id{D_0}}
	\ar[d]_{\id{D_0}}& D_{0} \ar@{>->}[d]^{f_0}&K_1 \ar@{>->}[d]_{l_1}\ar[r]^{k_{\der{E},0}}&
	D_{\der{E},0} \ar@{>->}[d]^{f_{\der{E},0}} & K_1 \ar[r]^{u}
	\ar@{>->}[d]_{l_1}& P \ar@{>->}[d]^{f_0\circ
		p_0}\\
	L_1\ar[r]^{i_1} \ar@/_.4cm/[rr]_{m_{\der{E},0}}&D_0
	\ar@{>->}[r]^{f_0}&G_0 & L_1\ar[r]_{m_{\der{E},0}}&G_0 &L_1
	\ar[r]_{m_{\der{E},0}}&G_0}\]

  	Since $f_0$ is a monomorphism, every square in the diagram on the left above
  	is a pullback. Thus the whole big square is a pullback too.
  	Applying \Cref{lem:radj} to the pair of squares on the right now gives
  	us the wanted $q_0\colon P\to D_{\der{E},0}$

@@

	Moreover, by \Cref{prop:pbpoad}, the first square below is a
	pullback, therefore, by \cref{rem:deco} the last three squares
	witness that $(j_0, j_1)$ is a strong independence pair too.
	\[
	\xymatrix@R=15pt{P\ar[r]^{q_1} \ar@{>->}[d]_{q_0}&Q_1
		\ar@{>->}[d]^{a_1}&K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1
		\ar[d]^{j_0} & K_0 \ar@{>->}[r]^{l_0} \ar[d]_{u_0}& L_0
		\ar[d]^{j_1}&K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0\ar[d]^{i_0} \\
		Q_ 0\ar[r]_{b_0}&H_1&P \ar[r]_{q_1} & Q_1 &P \ar[r]_{q_0} &
		Q_0&P \ar[r]_{p_1} & D_1}
	\]
\end{remark}

@@

\begin{proof}
	To fix the notation, let us depict the derivation $\der{D}=\{\dder{D}_i\}_{i=0}^2$:
	\[\xymatrix@C=13pt{L_0\ar[d]_{m_0}&& K_0 \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0 \ar@/^.35cm/[drrr]_(.4){i_0}|(.285)\hole \ar[dr]|(.28)\hole_{h_0} && L_1 \ar@/_.35cm/[dlll]^(.4){i_1} \ar[dl]|(.28)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[r]^{r_1} & R_1 \ar[dr]|(.28)\hole_{h_1} \ar@/^.35cm/[drrr]_(.4){j_0}|(.285)\hole  && L_2 \ar@/_.35cm/[dlll]^(.4){j_1} \ar[dl]|(.28)\hole^{m_2}& K_2 \ar[d]^{k_2}\ar@{>->}[l]_{l_2} \ar[rr]^{r_2} && R_2 \ar[d]_{h_2} \\G_0 && \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_{g_0}&& G_1  && \ar@{>->}[ll]^{f_1} D_1 \ar[rr]_{g_1}&& G_2 && \ar@{>->}[ll]^{f_2} D_2 \ar[rr]_{g_2}&& G_3 }\]

@@

	We begin switching $\dder{D}_0$ and $\dder{D}_1$ to get the following diagram.
	\[\xymatrix@C=15pt{L_1\ar[d]^{f_0\circ i_1}&& K_1 \ar[d]_{k'_1}\ar@{>->}[ll]_{l_1} \ar[r]^{r_1} & R_1 \ar@/^.35cm/[drrr]_(.4){a_0}|(.285)\hole \ar[dr]|(.28)\hole_{h'_1} && L_0 \ar@/_.35cm/[dlll]^(.4){a_1} \ar[dl]|(.28)\hole^{m'_0}& K_0 \ar[d]^{k'_0}\ar@{>->}[l]_{l_0} \ar[r]^{r_0} & R_0 \ar[dr]|(.28)\hole_{g_1\circ i_0} \ar@/^.35cm/[drrr]_(.4){b_0}|(.285)\hole  && L_2 \ar@/_.35cm/[dlll]^(.4){b_1} \ar[dl]|(.28)\hole^{m_2}& K_2 \ar[d]^{k_2}\ar@{>->}[l]_{l_2} \ar[rr]^{r_2} && R_2 \ar[d]_{h_2} \\G_0 && \ar@{>->}[ll]^{f'_0} D'_0 \ar[rr]_{g'_0}&& G'_1  && \ar@{>->}[ll]^{f'_1} D'_1 \ar[rr]_{g'_1}&& G_2 && \ar@{>->}[ll]^{f_2} D_2 \ar[rr]_{g_2}&& G_3 }\]
